---
title: "JAG1 Peptide Mapping and ICD Enrichment"
author: "Giulia Franciosa"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Overview

This analysis evaluates the peptide-level evidence for enrichment of the **Jagged-1 intracellular domain (JAG1-ICD)** in the RIME-DIA experiment.

Peptides assigned to JAG1 are classified according to whether their stripped peptide sequence maps within the JAG1 intracellular-domain sequence. The analysis compares the number and MS2 signal contribution of ICD-derived and non-ICD peptides and examines their abundance distributions.

The required Spectronaut peptide-level report is not stored in this GitHub repository. To reproduce the analysis, download the corresponding report from the PRIDE dataset and place it in the local `data/` directory.

# Setup

```{r setup}
library(tidyverse)
library(data.table)
library(here)

data_dir <- here("data")
plots_dir <- here("plots")

input_file <- file.path(
  data_dir,
  "20251113_134101_20220819_Report.tsv"
)
```

# 1. Load peptide-level Spectronaut data

The Spectronaut report is filtered to retain only peptide entries assigned to **JAG1**.

```{r load-data}
data <- fread(input_file) %>%
  as.data.frame()

data_jagged1 <- data %>%
  filter(PG.Genes == "JAG1")
```

# 2. Simplify run names

Run names are shortened to retain only the experimental group and replicate identifier.

```{r rename-runs}
data_jagged1 <- data_jagged1 %>%
  mutate(
    R.FileName = R.FileName %>%
      str_remove("^.*HPAF_") %>%
      str_replace("Sample15|Sample17", "R1") %>%
      str_replace("Sample16|Sample18", "R2") %>%
      str_replace("Sample21|Sample23", "R3") %>%
      str_replace("Sample22|Sample24", "R4")
  )


glimpse(data_jagged1)
```

# 3. Classify peptides as ICD or non-ICD

The amino-acid sequence of the JAG1 intracellular domain used in the analysis is defined below. Each JAG1 peptide is classified according to whether its stripped peptide sequence occurs within this sequence.

```{r classify-peptides}
jag1_icd_sequence <- paste0(
  "RKRRKPGSHTHSASEDNTTNNVREQLNQIKNPIEKHGANTVPIKDYENKNSKMSKIRTHNSEV",
  "EEDDMDKHQQKARFAKQPAYTLVDREEKPPNGTPTKHPNWTNKQDNRDLESAQSLNRMEYIV"
)

data_jagged1 <- data_jagged1 %>%
  mutate(
    ic = if_else(
      str_detect(jag1_icd_sequence, fixed(PEP.StrippedSequence)),
      "ICD",
      "non-ICD"
    )
  )

cat("Number of peptide observations:\n")
table(data_jagged1$ic)

cat("\nProportion of peptide observations:\n")
prop.table(table(data_jagged1$ic))
```

# 4. Overall peptide distribution

The relative number of peptide observations mapping to the ICD and non-ICD regions is summarized below.

```{r peptide-count-overall, fig.width=6, fig.height=5}
peptide_counts <- data_jagged1 %>%
  count(ic, name = "n")

p_peptide_count <- ggplot(
  peptide_counts,
  aes(x = "", y = n, fill = ic)
) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    color = "white"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Peptide region"
  ) +
  theme_void()

p_peptide_count

ggsave(
  file.path(plots_dir, "JAG1_peptide_count_ICD_vs_nonICD.png"),
  p_peptide_count,
  width = 6,
  height = 5,
  dpi = 300
)
```

# 5. Peptide distribution by run

The number of ICD and non-ICD peptide observations is shown for each MS run.

```{r peptide-count-run, fig.width=8, fig.height=5}
peptide_counts_run <- data_jagged1 %>%
  count(R.FileName, ic, name = "n")

p_peptide_run <- ggplot(
  peptide_counts_run,
  aes(x = R.FileName, y = n, fill = ic)
) +
  geom_col() +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 3
  ) +
  labs(
    x = "MS run",
    y = "Number of peptide observations",
    fill = "Peptide region"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_peptide_run

ggsave(
  file.path(plots_dir, "JAG1_peptide_count_by_run.png"),
  p_peptide_run,
  width = 8,
  height = 5,
  dpi = 300
)
```

# 6. Contribution to total JAG1 MS2 signal

The MS2 quantities of all JAG1 peptide observations are summed separately for ICD and non-ICD peptides. This estimates the relative contribution of each region to the total JAG1 peptide signal.

```{r ms2-signal, fig.width=6, fig.height=5}
signal_summary <- data_jagged1 %>%
  group_by(ic) %>%
  summarise(
    summed_MS2 = sum(PEP.MS2Quantity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    fraction = summed_MS2 / sum(summed_MS2)
  )

signal_summary
```

```{r ms2-signal-plot, fig.width=6, fig.height=5}
p_signal <- ggplot(
  signal_summary,
  aes(x = "", y = summed_MS2, fill = ic)
) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = scales::percent(fraction, accuracy = 1)),
    position = position_stack(vjust = 0.5),
    color = "white"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Peptide region"
  ) +
  theme_void()

p_signal

ggsave(
  file.path(plots_dir, "JAG1_MS2_signal_ICD_vs_nonICD.png"),
  p_signal,
  width = 6,
  height = 5,
  dpi = 300
)
```

# 7. Peptide-level intensity distribution

Because the same peptide sequence can be observed across multiple runs, a single representative intensity is calculated for each unique stripped peptide sequence using the median MS2 quantity across observations.

```{r peptide-median}
data_jagged1_median <- data_jagged1 %>%
  group_by(PEP.StrippedSequence) %>%
  summarise(
    median_MS2 = median(PEP.MS2Quantity, na.rm = TRUE),
    ic = first(ic),
    .groups = "drop"
  )
```

```{r peptide-intensity-plot, fig.width=6, fig.height=5}
p_intensity <- ggplot(
  data_jagged1_median,
  aes(x = ic, y = median_MS2)
) +
  geom_violin(trim = FALSE) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA
  ) +
  scale_y_log10(
    labels = scales::comma
  ) +
  labs(
    x = NULL,
    y = "Median peptide MS2 quantity"
  ) +
  theme_minimal(base_size = 12)

p_intensity

ggsave(
  file.path(plots_dir, "JAG1_peptide_median_MS2_intensity.png"),
  p_intensity,
  width = 6,
  height = 5,
  dpi = 300
)
```

# Session information

```{r session-info}
sessionInfo()
```
